---
title: "Recommendation System Report"
author: "Anil Nanayakkara"
date: "March 1, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

## R Markdown

Machine Learning Algorithm for Movie Recommendation System

```{r cars}
#############################################################
# Create edx set, validation set, and submission file
#############################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)

colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))
movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1)
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(ratings, movies, test_index, temp, movielens, removed)

# RMSE function
RMSE <- function(true_ratings, predicted_ratings) {
  sqrt(mean((true_ratings - predicted_ratings)^2))
}

# average rating among all movies
mu <- mean(edx$rating)

# regularization parameter
lamda <- 5

# regularized movie averages
movie_reg_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = sum(rating - mu)/(n() + lamda), n_i = n())

# regularized user averages
user_reg_avgs <- edx %>%
  left_join(movie_reg_avgs, by = "movieId") %>%
  group_by(userId) %>%
  summarize(b_u = sum(rating - mu - b_i)/(n() + lamda))

# movie genres separated into rows and split by individual genre
movie_data_listed_by_genre <- edx %>% 
  separate_rows(genres, sep = "\\|") %>%
  split(., .$genres)

rm(edx)

# genre effect of movie
genre_classified_movie_reg_avgs <- movie_data_listed_by_genre %>%
  lapply(function(x) {
    genre <- x$genres[1] 
    x %>% 
      left_join(movie_reg_avgs, by = "movieId") %>%
      left_join(user_reg_avgs, by = "userId") %>%
      group_by(movieId) %>% 
      summarize(b_ig = sum(rating - mu - b_i - b_u)/(n() + lamda), g = genre) 
  })

# genre effect of user
genre_classified_user_reg_avgs <- movie_data_listed_by_genre %>%
  lapply(function(x) {
    genre <- x$genres[1] 
    x %>% 
      left_join(movie_reg_avgs, by = "movieId") %>%
      left_join(user_reg_avgs, by = "userId") %>%
      left_join(genre_classified_movie_reg_avgs[[genre]], by = "movieId") %>%
      group_by(userId) %>%
      summarize(b_ug = sum(rating - mu - b_i - b_u - b_ig)/(n() + lamda), g = genre)
  })

# predict rating based on movieId, userId and genre
predicted_rating <- function(movie_id, user_id, genre) {
  # if the user has not rated any movies in this genre, then the genre effect for user is 0
  b_ug <- b_ug <- genre_classified_user_reg_avgs[[genre]] %>% 
    filter(userId == user_id) %>% .$b_ug
  if(length(b_ug) == 0) {
    b_ug <- 0
  }
  mu + b_ug +
    movie_reg_avgs %>% filter(movieId == movie_id) %>% .$b_i +
    user_reg_avgs %>% filter(userId == user_id) %>% .$b_u +
    genre_classified_movie_reg_avgs[[genre]] %>% 
      filter(movieId == movie_id) %>% .$b_ig
}

# prepare data and call function to predict  
predicted_ratings_split_by_genre <- function(userId, movieId, genres) {
  genre <- strsplit(genres, "\\|")[[1]][1]
  predicted_rating(movieId, userId, genre)
}

# apply vectors to predict functions
predicted_model_rating <- mapply(predicted_ratings_split_by_genre, 
                                 validation$userId,
                                 validation$movieId,
                                 validation$genres)

# calculated RMSE
rmse_calculated <- RMSE(validation$rating, predicted_model_rating)
```




EXECUTIVE SUMMARY


The goal of this project was to come up with a machine learning algorithm to predict what rating a user may give a movie based on movies he/she and other users have rated in the past. The main strategy was reducing the individual movie rating over the mean rating of all movies into different effects or biases.  These effects were then be used to predict movie and user combinations that have not been rated before.  


DATA SET


The data set included 10 million movie ratings by 69878 users for 10677 movies



OBSERVATIONS


(1) Movies that are better have a better rating that poor movies    which users do not like.  

(2) Different users rate the same movies differntly.

(3) Some movie genres get higher movie ratings than others.

(4) Users rate movies in some genres differently that other genres.



ANALYSIS


The approach taken to finding a solution to this problem was to calculate an average rating for all movies and then decompose the difference between the actual movie rating and the average for all all movies (mu) based on the observations on the data.

This resulted in obtaing the factors that contribute to the rating in the following way.

From observation (1) -> b_i = avg rating (for movie) - mu (movie effect)
                 (2) -> b_u = avg rating (for user) - mu - bi (user effect)
                 (3) -> b_ig = avg rating (for movie within genre) - b_i - b_u (movie effect for genre) 
                 (4) -> b_ug = avg rating (for user within genre) - b_i - b_u - b_ig (user effect for genre) 
  
These factors are then stored in separate lists and a prediction is made by adding up the factors:

prediction = mu + b_i + b_u + b_ig + b_ug


                
DATA CLEANSING AND EXPLORATION


The following strategies were followed to overcome problems inherent in the data.

1. movieId - was stored in a numeric vector, but some movieIds had a leading space. This made it difficult to vectorize the given dataframe by applying logic to rows of the data set.  To overcome this, vectors of movieId, userId and genres where extracted from the dataframe and applied as vectors using the mapply function.

2. genres - were stored as a list which needed to be separated. It was assumed that the same list of genres where applied to a particular movieId and in the same order as this is usually done in the data cleansing phase.

3. Movies that were rated very few times were rated very high or very low.  In order to penalize such cases, the algorithm regularized the rating with lamda value of 5 (since there are 4 effects that are being regularized a higher value was used)

RESULTS

The final Root Mean Square Error calculated was:
```{r}
rmse_results <- data_frame(method = "RMSE for Movie Recommendation Algorithm", RMSE = rmse_calculated)
rmse_results %>% knitr::kable()
```


VISUALIZATIONS


The following plots show how the effects used in the algorithm contributes to an accurate prediction.

```{r pressure, echo=FALSE}
# visualizations - Movie effect
movie_reg_avgs %>% ggplot(aes(x = b_i)) + geom_histogram() +
  labs(x = "Movie Effect") 

# visualizations - User effect
user_reg_avgs %>% ggplot(aes(x = b_u)) + geom_histogram() +
  labs(x = "User Effect") 

# functions to generate dataframes for visualizations
df <- function(x) {
  data.frame(genre = x[3], effect = x[2]) 
}

dfs <- function(y) {
  z <- data.frame()
  for (i in 1 : length(y)) {
    x <- df(y[[i]])
    z <- rbind(z, x)
  }
  z
}

# visualizations - Movie effect for individual movies within a genre
movie_genre_effs_df <- dfs(genre_classified_movie_reg_avgs)

movie_genre_effs_df %>% ggplot(aes(x = g, y = b_ig)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Genre", y = "Movie Effect for Genre") 

# visualizations - User effect for individual users within a genre
user_genre_effs_df <- dfs(genre_classified_user_reg_avgs)

user_genre_effs_df %>% ggplot(aes(x = g, y = b_ug)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Genre", y = "User Effect for Genre") 
```

CONCLUSION


The visualization shows that the effects used in the algorithm were significant in recommending movies based on prediction from those effects. There are diffences in movie effects for genres for individual movies as well as their means.  On the other hand even though the means do not change much for user effects for genres, the individual user effects across genres do vary more prominently.  The overall movie effect also shows a left skewed distribution where as the overall user effect is more normally distributed.

